Rapid technological progress in white light-emitting diodes \\ and its sources in innovation and technology spillovers

### 2023-03-14 2_SSL_EES/article/article_main.tex
minor edits

\begin{figure*}
 \centering
 \includegraphics[height=17cm]{2_SSL_EES/article/figures/chip_architecture_overview.pdf}
 \caption{Historical evolution of light-emitting diode chip architectures. Rows from top to bottom: classical chip with lateral current spreading; Osram’s thin GaN flip-chip (vertical) architecture; flip-chip architecture; chip-scale package flip-chip architecture. Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture. The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip. Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity. Years indicated correspond to the earliest identified patent priority date. Black frames around certain designs indicate chip designs not brought to large-scale production. Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well. Source: adapted and compiled from multiple patents and industry publications. See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
 \label{fgr:chip_architecture_overview}
\end{figure*}

@@

Historically, the progress in LED technology has commonly be described by pointing to impressive improvements in LED device performance, more specifically device brightness, electrical efficiency and manufacturing cost reductions \cite{Taki2019}. However,metrics related to consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, has also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000}\cite{Sandahl2006}\cite{CAIRD2008}\cite{murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition}\cite{cho2017white}. Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in: 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs. Next, we introduce and discuss the metrics that we use to track progress in each of these three areas.

@@

During the years following the market introduction of white LEDs, the primary metric of progress in solid-state lighting was typically luminous flux. This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes. In 2000, highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm)\cite{haitz2011solid}. Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case}\cite{haitz2011solid}. Today, however, highest performing devices yield in excess of 1600lm, the equivalent of a 100W incandescent bulb \cite{cree2020bright}. LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead higher efficiency and quality of light instead of brightness. Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, by now, it fails to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}. Specifically,  the stagnating levels of luminous flux in highest performing devices fail to capture the majority of research, which serves to improve electrical efficiency at constant brightness.

Were therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe the different physical loss channels within the device to describe progress in light-emitting diode technology:

@@

(Efficiency) droop describes the decrease of device internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}. This is often treated separately from internal quantum efficiency in literature due to its importance in high-power devices. It is defined as

@@

Thereby, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.

@@

(Light) conversion efficiency (LCE) describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs. These losses are the sum of the Stokes loss and well as scattering/absorption losses. It is defined as

@@

where $E_i$ denotes the total energy of light at the color corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}. Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all components (\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen).

@@

The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}. Any metric relevant to customer experience can thus be calculated from the spectrum alone. The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra  of the down-conversion phosphor used in the device. It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices. Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: Colour Rendering Index (CRI) and Colour Temperature. We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}. This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}. It is thus beyond the scope of this work.

@@

The Colour Rendering Index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}. The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}. CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}. However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de-facto industry standard for describing colour fidelity of light sources \cite{DOE2016LED}. High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}. It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}. However, these niche applications remain outside our focus on general illumination. Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.

@@

According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change  in a single sub-efficiency variable a between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}

@@

To document historical improvements in LED device performance accurately, we need data on all sub-efficiencies for the selected device architectures and periods covered. However, the scope of data reporting in scientific literature and industry publications is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance. For this reason, our data collection efforts were supplemented by performance calculations for individual sub-efficiencies where possible and necessary. Of the device sub-efficiencies, those related to the emission spectrum  were computed from the spectral data often reported in LED device specifications. In particular, we used the \texttt{colour-science} package for Python\cite{python-colour} to calculate the luminous efficacy of radiation, colour rendering performance and luminous efficacy of radiation of phosphor down-conversion of blue light on the basis of available LED emission spectra. This approach allowed us to fully quantify the improvements related to phosphor development in LEDs.

@@

We address the limitations of data availability for the LED manufacturing cost by conducting bottom-up process-step resolution LED manufacturing cost modelling. In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.

@@

Given this situation, we developed our own bottom-up LED manufacturing cost model. The model structure is generally based on the 2012 LEDCOM cost model, but we expand it significantly both in scope and in its ability to capture historical trends. The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts. For each of these three periods, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the \textit{Internet Archive}, and expert interviews. Because the architecture of LED chips has changed significantly since the introduction of the first commercial devices in 1996, three different chip architectures were initially considered in the model: classical chips, flip chips, and chip-scale package flip chips. The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents. In addition, two LED life cycle analyses \cite{scholand2012life}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs. These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.

The aggregate result of the cost model is the manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory. Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered. We also note that the purpose of our cost model is not to give specific estimates of LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume. It instead assumes an ideal factory with an assumed location in the United States and associated overhead costs related to the operation of the factory. It also assumes the use of the most up-to-date equipment for that year. Even with these simplifying assumptions, the model reasonably identifies the impact that changes in single process steps can have on the total LED manufacturing cost. An important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we were able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low- to mid-power devices. Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.

@@

In Figure \ref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of corresponding loss channels. We find that there has been consistent progress across all device sub-efficiencies in the recorded period. Specifically, between 2003 and 2020, forward voltage efficiency increased from 70\% to 99.5\%, internal quantum efficiency from 55\% to 90\%, electrical droop from 65\% to 90\%, light extraction efficiency from 60\% to 90\%, spectral efficiency from 74\% to 83\%, conversion efficiency (red) from 11\% to 45\%, conversion efficiency (green) from 19\% to 61\%. Notably, some sub-efficiencies for the most recent devices considered in our study are now within $\sim10\%$ of their respective physical limits. The exception is spectral efficiency, which at $\sim17\%$ below the physical limit shows larger potential for further improvements.

@@

\begin{figure*}[h!]
 \centering
 \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_efficiency.pdf}
 \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA. Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study. Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for the four years used in Figure \ref{fgr:waterfall}: 2003, 2010, 2016, and 2020. To provide additional historical context, data for 1997 is included for selected sub-efficiencies. "N/A" denotes sub-efficiencies where 1997 data could not be calculated to different reasons: $\eta_{Vf}$, $\eta_{droop}$ depend on the device current, which was below 350mA in 1997, making comparison with contemporary devices difficult. $\eta_{C(R)}$,$\eta_{C(Y/G)}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997. Note that indium tin oxide (ITO) affects different sub-efficiencies depending on the chip architecture in question. For instance, in modern flip-chip architectures, light-extraction efficiency no longer depends on ITO, as can be seen from Figure \ref{fig:chip_architecture_overview}. Physical limits on sub-efficiencies are indicated by black horizontal lines. The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column. $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_Droop$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C(R)}$ - conversion efficiency for red phosphors; $\eta_{C(Y/G)}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table \ref{tab:phosphors}.  Note that physical limit on $\eta_{V_f}$ is above 100\% due to quantum effects which depends on electrical device parameters \cite{david2016electrical}. Source: own elaboration based on data in Figure \ref{fgr:waterfall} and Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}.}
 \label{fgr:breakthroughs_efficiency}
\end{figure*}

@@

\begin{table*}[h!]
    \small
    \centering
    \begin{tabularx}{\textwidth}{|l|l|l|X|X|l|}
    \hline
        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{SI Sec.} \\ \hline
        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3 \\ \hline
        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 3.1 \\ \hline
        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3 \\ \hline
        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3 \\ \hline
        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3 \\ \hline
        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3 \\ \hline
        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3 \\ \hline
    \end{tabularx}
    \caption{LED down-conversion materials (phosphors and quantum dots) related to improvements in consumer experience metrics identified in this study. The \textit{year} column represents the earliest reported application of invention in white LEDs. These differ from the years used in Figure \ref{fgr:consumer_experience} that correspond to the earliest publication of spectral data for representative LED products that relied on those materials. The \textit{SI Sec.} column refers to the corresponding section in the Supplementary Information that contains references and a detailed description of the history of corresponding invention. Abbreviations: Desig. - Designation, Sec. - Section.}
    \label{tab:phosphors}
\end{table*}

@@

Results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages are presented in Figure \ref{fgr:costmodel}. We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.

Among the factors contributing to the LED cost reductions over time, improved manufacturing yields and increases in the wafer size rather than particular LED innovations are found to be responsible for the largest contribution to the overall cost reduction. In the case of manufacturing yield, the higher it is, the less inputs are wasted on the production of a single LED package. With the total manufacturing yield dramatically improving from $\sim25\%$ in 2003 to $\sim75\%$ in 2020 (compare Figure \ref{fgr:costmodel}, Panels A-C), it is not surprising that the total yielded LED manufacturing cost significantly declined over this period. In the case of wafer diameter, the larger the wafer, the more LED chips can be produced from a single wafer. The wafer diameter commonly used in LED manufacturing has been steadily increasing 2003 from 51mm (2 inch) in 2003 to 200mm ($\sim$8 inches, referred to as “8 inch") in 2020. We capture this in the model by assuming the following wafer diameters and calculating the associated number of die per wafer (DPW) \cite{de2005investigation}: $d(2003)\sim 50 mm \rightarrow851 DPW$, $d(2020)\sim200 mm \rightarrow 26,838 DPW$. With more than thirty-fold increase in the number of die per wafer between 2003 and 2020, the contribution of the whole-of-wafer processing steps to the total cost of manufacturing an individual LED chip and package has dramatically declined over time. As the number of die per wafer increases, the packaging steps, which in the classic chip architecture must be performed separately for each individual LED chip, carry a significantly larger share of the total cost in 2020 than in 2003 (compare Panels A-C in Figure \ref{fgr:costmodel}). However, as our interviewees noted, while growing LEDs on larger wafers is economically desirable, it is associated with engineering and epitaxy challenges as well as high up-front cost of new equipment.

Our findings are further supported by apreliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the model to variation in its main parameters decreases over time with the increase of the number of DPW. We also provide a comparison of our model with past cost calculations and projections published by the US DOE on the basis of the LEDCOM cost model\cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables\cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} in Section 2 of the Supplementary Information.

>>>>

\begin{figure*}
 \centering
 \includegraphics[height=17cm]{2_SSL_EES/article/figures/chip_architecture_overview.pdf}
 \caption{Historical evolution of light-emitting diode chip architectures. Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture. Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture. The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip. Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity. Years indicated correspond to the earliest identified patent priority date. Black frames around certain designs indicate chip designs not brought to large-scale production. Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well. Source: adapted and compiled from multiple patents and industry publications. See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
 \label{fgr:chip_architecture_overview}
\end{figure*}

@@

Historically, the progress in LED technology has commonly be described by pointing to impressive improvements in LED device performance, more specifically device brightness, electrical efficiency and manufacturing cost reductions \cite{Taki2019}. However, metrics related to consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000}\cite{Sandahl2006}\cite{CAIRD2008}\cite{murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition}\cite{cho2017white}. Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in: 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs. Next, we introduce and discuss the metrics that we use to track progress in each of these three areas.

@@

During the years following the market introduction of white LEDs, the primary metric of progress in solid-state lighting was typically luminous flux. This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes. In 2000, the highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm)\cite{haitz2011solid}. Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case}\cite{haitz2011solid}. Today, however, the devices with the highest performance yield in excess of 1600lm, the equivalent of a 100W incandescent bulb \cite{cree2020bright}. LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead on higher efficiency and quality of light instead of brightness. Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}. Specifically,  the stagnating levels of luminous flux in highest performing devices does not capture a major research focus, which is improving electrical efficiency at constant brightness.

We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe the different physical loss channels within the device to describe progress in light-emitting diode technology.

@@

Efficiency droop describes the decrease of device internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}. This is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices. It is defined as

@@

Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.

@@

Light conversion efficiency (LCE) describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs. These losses are the sum of the Stokes loss and well as scattering/absorption losses. It is defined as

@@

where $E_i$ denotes the total energy of light at the color corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}. Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all components, i.e. \textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen.

@@

The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}. Any metric relevant to customer experience can thus be calculated from the spectrum alone. The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra  of the down-conversion phosphor used in the device. It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices. Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: Colour Rendering Index (CRI) and Colour Temperature. We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}. This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}. As a result, it is beyond the scope of this work.

@@

The Colour Rendering Index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}. The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}. CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}. However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DOE2016LED}. High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}. It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}. However, these niche applications remain outside our focus on general illumination. Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.

@@

According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change in a single sub-efficiency variable $a$ between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}

@@

To document historical improvements in LED device performance accurately, we need data on all sub-efficiencies for the selected device architectures and periods covered. However, the scope of data reporting in scientific literature and industry publications is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance. For this reason, our data collection efforts were supplemented by performance calculations for individual sub-efficiencies where possible and necessary. Of the device sub-efficiencies, those related to the emission spectrum  were computed from the spectral data often reported in LED device specifications. In particular, we used the \texttt{colour-science} package for Python\cite{python-colour} to calculate the luminous efficacy of radiation, colour rendering performance and luminous efficacy of radiation of phosphor down-conversion of blue light on the basis of available LED emission spectra. This approach allowed us to  quantify the improvements related to phosphor development in LEDs.

@@

We address the limitations of data availability for the LED manufacturing cost by developing and applying a bottom-up LED manufacturing cost model with process-step resolution. In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.

@@

Given this situation, we developed our own bottom-up LED manufacturing cost model. The model structure is generally based on the 2012 LEDCOM cost model, but we expand it significantly both in scope and in its ability to capture historical trends. The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts. For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the \textit{Internet Archive}, and expert interviews. Because the architecture of LED chips has changed significantly since the introduction of the first commercial devices in 1996, three different chip architectures were initially considered in the model: classical chips, flip chips, and chip-scale package flip chips. The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents. In addition, two LED life cycle analyses \cite{scholand2012life}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs. These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.

The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory. Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered. We also note that the purpose of our cost model is not to give specific estimates of LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume. It instead assumes an hypothetical factory with an assumed location in the United States and associated overhead costs related to the operation of the factory. It also assumes the use of the most up-to-date equipment for that year. Even with these simplifying assumptions, the model reasonably identifies the impact that changes in single process steps can have on the total LED manufacturing cost. An important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we were able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low- to mid-power devices. Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.

@@

In Figure \ref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels. We find that there has been consistent progress across all device sub-efficiencies in the recorded period. Specifically, between 2003 and 2020, forward voltage efficiency increased from 70\% to 99.5\%, internal quantum efficiency from 55\% to 90\%, electrical droop from 65\% to 90\%, light extraction efficiency from 60\% to 90\%, spectral efficiency from 74\% to 83\%, conversion efficiency (red) from 11\% to 45\%, conversion efficiency (green) from 19\% to 61\%. Notably, some sub-efficiencies for the most recent devices considered in our study are now within $\sim10\%$ of their respective physical limits. The exception is spectral efficiency which, at $\sim17\%$ below the physical limit, shows larger potential for further improvements.

@@

\begin{figure*}[h!]
 \centering
 \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_efficiency.pdf}
 \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA. Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study. Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for the four years used in Figure \ref{fgr:waterfall}: 2003, 2010, 2016, and 2020. To provide additional historical context, data for 1997 is included for selected sub-efficiencies. "N/A" denotes sub-efficiencies where 1997 data could not be calculated to different reasons: $\eta_{Vf}$, $\eta_{droop}$ depend on the device current, which was below 350mA in 1997, making comparison with contemporary devices difficult. $\eta_{C(R)}$,$\eta_{C(Y/G)}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997. Note that indium tin oxide (ITO) affects different sub-efficiencies depending on the chip architecture in question. For instance, in modern flip-chip architectures, light-extraction efficiency no longer depends on ITO, as can be seen from Figure \ref{fgr:chip_architecture_overview}. Physical limits on sub-efficiencies are indicated by black horizontal lines. The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column. $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_Droop$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C(R)}$ - conversion efficiency for red phosphors; $\eta_{C(Y/G)}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table \ref{tab:phosphors}.  Note that physical limit on $\eta_{V_f}$ is above 100\% due to quantum effects which depends on electrical device parameters \cite{david2016electrical}. Source: own elaboration based on data in Figure \ref{fgr:waterfall} and Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}.}
 \label{fgr:breakthroughs_efficiency}
\end{figure*}

@@

\begin{table*}[h!]
    \small
    \centering
    \begin{tabularx}{\textwidth}{|l|l|l|X|X|l|}
    \hline
        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{SI Sec.} \\ \hline
        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3 \\ \hline
        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 3.1 \\ \hline
        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3 \\ \hline
        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3 \\ \hline
        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3 \\ \hline
        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3 \\ \hline
        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3 \\ \hline
    \end{tabularx}
    \caption{LED down-conversion materials (phosphors and quantum dots) related to improvements in consumer experience metrics identified in this study. The \textit{year} column represents the earliest reported application of invention in white LEDs. These differ from the years used in Figure \ref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for representative LED products that relied on those materials. The \textit{SI Sec.} column refers to the corresponding section in the Supplementary Information that contains references and a detailed description of the history of corresponding invention. Abbreviations: Desig. - Designation, Sec. - Section.}
    \label{tab:phosphors}
\end{table*}

@@

Figure \ref{fgr:costmodel} shows key results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages. We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.

Among the factors contributing to the LED cost reductions over time, improved manufacturing yields and increases in the wafer size rather than particular LED innovations are found to be responsible for the largest contribution to the overall cost reduction. In the case of manufacturing yield, the higher it is, the less inputs are wasted on the production of a single LED package. With the total manufacturing yield dramatically improving from $\sim25\%$ in 2003 to $\sim75\%$ in 2020 (compare Figure \ref{fgr:costmodel}, Panels A-C), it is not surprising that the total yielded LED manufacturing cost significantly declined over this period. In the case of wafer diameter, the larger the wafer, the more LED chips can be produced from a single wafer. The wafer diameter commonly used in LED manufacturing has been steadily increasing 2003 from 51mm (2 inch) in 2003 to 200mm ($\sim$8 inches, referred to as “8 inch") in 2020. We capture this in the model by assuming the following wafer diameters and calculating the associated number of die per wafer (DPW) \cite{de2005investigation}: $d(2003)\sim 50 mm \rightarrow851 DPW$, $d(2020)\sim200 mm \rightarrow 26,838 DPW$. With more than a thirty-fold increase in the number of die per wafer between 2003 and 2020, the contribution of the whole-of-wafer processing steps to the total cost of manufacturing an individual LED chip and package has dramatically declined over time. As the number of die per wafer increases, the packaging steps, which in the classic chip architecture must be performed separately for each individual LED chip, carry a significantly larger share of the total cost in 2020 than in 2003 (compare Panels A-C in Figure \ref{fgr:costmodel}). However, as our interviewees noted, while growing LEDs on larger wafers is economically desirable, it is associated with engineering and epitaxy challenges as well as high up-front cost of new equipment.

Our findings are further supported by a preliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW. We also provide a comparison of our model with past cost calculations and projections published by the US DOE on the basis of the LEDCOM cost model\cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables\cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} in Section 2 of the Supplementary Information.
